Forecasting climate change impacts on plant population dynamics at large spatial extents: a test case with sagebrush (\emph{Artemisia}) species

### 2015-11-03 docs/ms_sageAbundance.Rmd
updates author list
updates climate projectsion for just 18 models with all three RCP scenarios

\textsc{\small{Andrew T. Tredennick\footnote{Corresponding author: atredenn@gmail.com}\textsuperscript{1}, Mevin B. Hooten\textsuperscript{2,3,4}, Cameron L. Aldridge\textsuperscript{5}, Collin G. Homer\textsuperscript{6}, David T. Iles\textsuperscript{1}, Andrew Kleinhesselink\textsuperscript{1}, Eric LaMalfa\textsuperscript{1}, Rebecca Mann\textsuperscript{1}, and Peter B. Adler\textsuperscript{1}}}

@@

Abstract
--------
Population modeling inference has traditionally been limited to small scale observational studies.
Yet, global environmental changes, like climate change, tend to play out at the landscape and regional scales, creating a mismatch between current tools and current problems.
New technology is allowing for data products on massive scales covering entire ecosystems and teams are post-processing these data products to turn them into useful ecological information.
For example, plant ecologists now have access to species-specific estimates of plant percent cover through time at spatial extents ranging from landscapes to continents.
The real strength to be gained using this new form of ecological information will be to move beyond simple descriptive summaries of the data, and couple them with sophisticated model-based statistical approaches that can accommodate more realistic ecological mechanisms.
A major hurdle in properly analyzing these types of data is in building and fitting statistical models that have valid assumptions.
Rigorously accounting for spatial autocorrelation has been a crux in former analyses at large scales, but we can now fit appropriate models using modern, computationally efficient model reparameterizations and dimension reduction techniques.

Here we advance spatio-temporal modeling of plant populations by applying recent developments in the modeling of animal abundances to remotely-sensed estimates of plant percent cover at the landscape scale.
We use a 27 year time series of sagebrush (*Artemisia* spp.) percent cover over a $2.5\times5$ km landscape in southwestern Wyoming to fit a spatio-temporal population model that includes climate covariates.
We then use the model to forecast future population states (percent cover) under projected climate scenarios.
Our analysis suggests sagebrush will increase in cover at this high elevation site, primarily driven by a positive response to previous year's precipitation.
However, gains in cover are not homogeneous in space.
Such spatially explicit estimates of increases/decreases in plant abundance have, until now, been elusive.
The spatiotemporal model we describe can be applied to any spatially explicit time series of plant cover.
As such datasets become increasingly available and precise, we expect the methods we use here to become an appealing alternative to species distribution models that lack population dynamics and, generally, estimates of species abundance.

@@

Introduction
-------------
Forecasting the impacts of climate change on plant populations and communities is a central challenge of modern ecology [@Clark2001; @Petchey2015].
Population models are ideally suited for meeting such a challenge because they allow inference on population abundance and dynamics (year-to-year changes in abundance) in relation to climate drivers.
However, inference from population models is typically limited to small spatial extents because the data required is difficult to collect across wide species ranges.
Indeed, almost every study of plant population dynamics relies on demographic observations recorded at the meter to sub-meter scale [see, e.g., @Salguero-Gomez2015].
Local-scale demographic data make building population projection models an easy task [@Ellner2006; @Rees2009; @Adler2012], but it is very difficult to extrapolate small-scale studies to large spatial extents with any certainty because the data likely only represent a small subset of parameter space and environmental conditions [@Freckleton2011; @Queenborough2011].
Thus, the real challenge is not to simply make population forecasts, but to do so at spatial scales relevant to policy and management decisions [@Queenborough2011, Tredennick and Adler in review].

Species distribution models (SDMs), on the other hand, are easily applied at landscape to continental scales [e.g., @Maiorano2013; @Clark2014].
In a climate change context, SDMs typically rely on static associations between contemporary climate and a species' distribution or, more rarely, abundance to project future distribution or abundance [@Elith2009].
Thus, interannual dynamics that reflect population responses to yearly climate variations are not represented in SDMs.
Long-term trends in species' occurence can be detected and correlated with climate variables, but these may or may not be representative of a species' sensitivity to climate.
Likewise, since SDMs typically rely on occurence data, their projections of habitat suitability or probability of occurence provide little information on the future states of populations in the core of their range -- in areas where a species currently exists and is expected to persist in the future.
Despite these limitations, SDMs remain one of the few ways to make forecasts of species responses to climate change at large spatial extents.

It seems the ideal tool for forecasting future population states under climate change is a large-scale dynamic population model [@Schurr2012a; @Merow2014].
Unfortunately, such methods have been elusive due to the lack of time series data at large spatial extents and the statistical machinary to fit high-dimensional spatial models.
Fortunately, two recent technological advances put us in the position to develop large-scale population models.
First, remotely-sensed (RS) time series of species-specific plant cover can provide data on interannual population dynamics that can be correlated with yearly climate fluctuations at large spatial extents.
RS time series have revolutionized models of ecosystem-level processes [e.g., @Running2004] and have been used to detect long term trends in plant population abundance [e.g., @Homer2015], but they have yet to be used to drive a dynamic population model.
Second, animal population modelers have developed efficient dimension reduction and reparameterization techniques to fit high-dimension spatiotemporal models [see @Conn2015 for a review]. 
However, these methods have yet to be use in a dynamic setting.

Building on these technological advances, we develop a statistical approach for modeling plant populations over large spatial extents using a process model motivated by Gompertz density-dependent growth.
Our approach leverages a unique remotely-sensed time series of *Artemisia* (sagebrush) species [@Homer2012] to fit a spatio-temporal model of plant population dynamics at the landscape scale.
We use advanced dimension reduction techniques borrowed from the animal modeling literature [@Conn2015] to rigorously account for spatial autocorrelation and to produce spatially-explicit estimates of sagebrush percent cover.
We then use the model to forecast the impacts of climate change on sagebrush percent cover.

Sagebrush ecosystems offer an ideal testing ground for our new modeling approach.
Their distribution is widespread, sagebrush species are sensitive to climate, and forecasts of future sagebrush ecosystems are in high demand due to the precarious status of sage-grouse (*Centrocercus* spp.).
Nearly all species distribution models predict that much of the area occupied by sagebrush ecosystems today will become unsuitable for sagebrush due to climate change, resulting in a dramatic loss overall in the extent of sagebrush habitat by the end of this century [@Shafer2001; @Neilson2005; @Bradley2010; @Schlaepfer2012; @Still2015].
Sagebrush is thought to depend on winter snow and rain that recharges deeper soil layers [@Pechanec1937; @Schlaepfer2011; @Germino2014].
Models suggest that warming could drive sagebrush losses at low elevations and southern parts of its distribution by making snowmelt earlier and increasing evaporation, leading to soil water deficits in the spring [@Schlaepfer2012; @Schlaepfer2014].
In contrast, at higher elevations and colder regions, warming and earlier snowmelt could lengthen the growing season and increase sagebrush occurrence [@Schlaepfer2012; @Schlaepfer2014].
Thus, the response of sagebrush to climate will likely vary in space depending on elevation, among other factors.
Direct observations of individual plants and cover on experimental plots tend to agree with these models:  growth tends to respond negatively to spring and summer temperatures [@Miglia2005; @Poore2009; @Apodaca2013] except at higher elevations where earlier snowmelt may allow for a longer growing season and an increase in sagebrush growth and cover [@Perfors2003; @Harte2015].

@@

Materials and Methods
---------------------
### Data
#### Remotely-sensed time series
To demonstrate our modeling approach we use a subset of a remotely-sensed time series of sagebrush (\emph{Artemisia} spp.) shrub percent cover in Wyoming[^link0] [@Homer2012].
As part of a separate study, sagebrush percent cover was estimated using a regression tree to relate ground reflectances retrieved by three sources of optical imagery (QuickBird, Landsat, and AWiFS) to 1,780 field observations of sagebrush cover distributed across Wyoming.
The regression tree model was further validated using another 500 field observations. For Wyoming sagebrush, the model achieved an $R^{2}$ = 0.65.
To hind-cast sagebrush cover the regression tree model was applied to historical remote sensing images. 
This resulted in an annual time series of sagebrush cover at 30 meter resolution from 1984 to 2011.
In this dataset, values represent the percentage of a 30 $\times$ 30 meter pixel covered by sagebrush.
In our study, we focused on a 5,070 $\times$ 2,430 meter subset totaling 13,689 30 $\times$ 30 meter pixels each year (Figure 1).
Thus, the full dataset contains 369,603 observations (27 years $\times$ 13,689 pixels).

@@

To forecast future population states, we required future projections of climate for our study area.
We used the most recent climate projections used by the Intergovernmental Panel on Climate Change (IPCC), the Coupled Model Intercomparison Project 5 (CMIP5; *available online*)[^link2].
The CMIP5 provides projections from a suite of global circulations models (Table X1).
To simulate equilibrium sagebrush cover under projected future climate (see below), we averaged all model projections for the centroid of our study area over the time period of 2050 to 2100.
We did the same for the time period of 1950 to 2000 so we could calculate projected changes in climate.
To keep the temporal variability from observed climate consistent for climate change scenarios, we adjusted observed climate according to projected changes.
For temperature, we subtracted the average historic (1950-2000) temperature from the average projected (2050-2100) temperature. 
For precipitation, we calculated the proportional change: we subtracted average projected (2050-2100) precipitation from average historic (1950-2000) precipitation, and then divided by average historic (1950-2000) precipitation. 
We calculate precipitation and temperature changes for three ``Representative Concentration Pathways'': RCP 4.5, RCP 6.0, and RCP 8.5 (*described online*)[^link3].
The three RCPs correspond to stabilization of radiative forcing before 2100, after 2100, and increasing greenhouse gas emissions, respectively.
We implemented projected temperature and precipitation changes in the model by perturbing observed annual values by the values in Table 1.

@@

In another set of simulations (see below), we used yearly GCM projections from 2012 to 2098 to make annual forecasts starting from the end point of the remotely sensed sagebrush cover data (ends in 2011). We aggregated daily GCM output for each model and RCP scenario into the yearly climate covariates used to fit our model.

@@

\vspace{-4em}
\begin{align}
\text{log}(\mu_{i,t}) &= \underbrace{\beta_{0,t} + \beta_{1}y_{i,t-1}}_\text{temporal} + \underbrace{\textbf{x}_{t}'\boldsymbol{\gamma}}_\text{climate} + \underbrace{\eta_{i}}_\text{spatial}.
\end{align}
\vspace{-3em}

Our model of percent cover change includes a temporal effect of log-transformed cover in the previous year ($y_{i,t-1}$), climate effects ($\textbf{x}_t$), and a spatial random effect ($\eta$) for each pixel *i*.
Climate effects were standardized [$(x_{i}-\bar{x}) / \sigma(x)$] to improve convergence during the model fitting stage and to allow for easier prior specification.
The intercept, $\beta_{0,t}$, was allowed to vary through time so that climate variables were not over-fit.
We used a Poisson likelihood because integer percent cover values in the sagebrush data product can be considered a form of count data.
We also evaluated a negative binomial model, but found little evidence for overdispersion beyond what our model was already accomodating via the spatial random effects ($\boldsymbol{\eta}$). There was no evidence of zero-inflation in our data, but see below (*Accomodating zeros*) for how we dealt with the small number of zero percent cover observations.

Fitting the model with a spatial random effect ($\eta$) is computationally demanding when a large number of observations are considered.
This is due to the required calculations of the resulting covariance matrices, which increase as a cubic function of the number of locations [@Wikle2010].
To overcome these computational constraints, we employed a dimension reduction strategy to reduce the number of parameters that need to be estimated to account for spatial variation.
Reducing the number of effective spatial random effect parameters speeds up estimation by reducing the size of the covariance matrices that need to be inverted at each MCMC iteration.
Fitting models that appropriately account for spatial autocorrelation over large spatial extents would be virtually impossible without these modern techniques.
We follow a common dimension reduction strategy, and express the high dimensional spatial random effect, $\boldsymbol{\eta}$, as the product of an expansion matrix, \textbf{K}, and a smaller parameter vector, $\boldsymbol{\alpha}$ [e.g., @Hooten2003; @Hooten2007a; @Conn2015].
We can then define the spatial effect as

@@

\vspace{-4em}
\begin{align}
w_{s,m} &= \text{exp}\left(\frac{-d_{s,m}}{\sigma}\right).
\end{align}
\vspace{-4em}

where $d_{s,m}$ is the Euclidean distance between the centroid of sample cell \emph{s} and the location of knot \emph{m}, and $\sigma$ is the kernel bandwidth.
It is possible through exhaustive model selection and fitting to determine the optimal form of the kernel and to estimate optimal values for $\sigma$ [@Higdon2002; @Hooten2015].
However, given the relative size of our dataset and computational limitations, we defined kernels around 121 knots (Fig. X#) whose nearest neighbor distances are approximately equal to the range of spatial dependence in residuals from a simple GLM fit without climate covariates and the spatial random effect (Appendix X). The Bayesian posterior distribution can be expressed as

@@

### Computing
####Fitting the model
We fit the model in R [@R2013] using the 'No-U-Turn' Hamiltonian Monte Carlo sampler in Stan [@stan2014] and the RStan package [@rstan2014].
We obtained posterior distributions of all model parmaters from three MCMC chains comprised of 1,000 iterations each, after discarding an initial 1,000 iterations as burn in.
Such short MCMC chains may seem unusual to those more familiar with other MCMC samplers (e.g., JAGS or WinBUGS).
However, the Stan algorithm is extremely efficient, meaning that, compared to other samplers, fewer iterations are required to achieve convergence.
Each MCMC chain was initialized with unique parameter values and the model was fit in parallel using the Utah State University High-Performance Computing facility.
Model fitting required five days.
We assessed convergence visually and calculated scale reduction factors [< 0.01, @Gelman1992; @Gelman2009].

####Simulating the process
We performed four sets of simulations: (1) to compare observed and simulated equilibrium cover, (2) to compare observed and simulated year- and location-specific cover, (3) to forecast future equilibrium population states under projected climate change, and (4) to make annual forecasts of sagebrush cover starting the final year of our observations and ending in year 2098. 
Using the posterior distribution of model parameters, we simulated a matrix of pixels equal to the size of the study area (13,689 pixels or matrix elements).
For simulations (1) and (3) we initialized all pixels with arbitrarily low cover (1%) and then projected the model forward by randomly drawing climate covariates from the observed climate time series (for 1) or a pertubed climate time series (for 3).
We ran equilibrium simulations (1 and 3) for 600 time steps and then compared the output across simulations, after discarding and initial 100 time steps.
For simulation (2), we initialized each pixel with its actual value for time *t* and cell *s* and projected the model forward one time step and compared the one-step forecast with the observed value.
For simulation (4), we initialized each pixel with the final observed value in 2011 and then projected the model forward based on GCM yearly weather projections.
We used the posterior mean of each parameter for all simulations, except for (4) where we ran 50 simulations with unique sets of parameters from the MCMC chains.
The random year effects were not used in any simulation.

@@

When we simulated the pixel-based population model based on observed climate it was able to reproduce average observed percent cover in space (Fig. 4A,B).
Our model shows a tendency to underpredict perennially-low percent cover pixels (Fig. 4C), but does a better job at predicting high cover pixels.
Nonetheless, observations and predictions are qualitatively similar (Fig. 4A, B).
Point predictions are most confident, though slightly biased, in low percent cover pixels (Fig. 4D).
The model is also able to adequately reproduce observed dynamics when we make one-step-ahead predictions based on observed climate and cover in the previous year for each pixel.
When we made these in-sample, one-step-ahead forecasts, the model achieved an RMSE = 4.31, in units of percent cover.
The correlation between observations and predictions is 0.62.

When we apply the fitted model to IPCC climate change scenarios, the model predicts gains in sagebrush percent cover, on average (Figs. 5, 6).
The spatial effect remains strong enough in low cover regions to counteract the amplifying effect of projected precipitation increases (Fig. 5).
Thus, our model predicts an increase in sagebrush cover heterogeneity because we do not lose low cover pixels and gain higher cover pixels (Fig. 5).
For the annual forecasts, we show spatially-averaged values and the associated uncertainty due to variability in GCM projections, variability in model parameters, and uncertainty in our process model (Fig. 6).
Based on our model and GCM projections, we forecast an average increase in sagebrush cover at our study area, but a decrease is not outside the realm of possibility (shaded regions in Fig. 6).
This is expected given the positive effect of precipitation on sagebrush cover change (Fig. 3).

Discussion
-----------
Landscape-scale population models for plant species have been elusive and challenging to implement [@Schurr2012a; @Merow2014].
Yet, the need to forecast population responses to climate change over large spatial extents is clear, as demonstrated by the wide application of species distribution models [see @Clark2014].
We introduced a new statistical approach to modeling plant populations that overcame two major hurdles for spatially-explicit population models.
First, we used species-specific remote sensing data as a response variable.
Thus, we had a spatially-continuous data product of sagebrush abundance over a large spatial extent, meaning we could fit a model at the same spatial resolution and extent as the data. 
In so doing we, overcame a major limitation of most population models that are fit using detailed data from a single location.
Second, borrowing from the vast literature on spatio-temporal modeling of animal abundance [e.g., @Conn2015], we employed a dimension reduction strategy to rigorously account for spatial autocorrelation.
Accounting for spatial autocorrelation allows for statistically rigorous inference.
While important statistically, in terms of ecological inference, we can be confident in our estimates of climate effects on sagebrush cover change in our study region.
It also provided a way to obtain spatially-explicit predictions at a resolution below that of the climate covariates (Figs. 4,5).

The amount of precipitation in the previous year (*pptLag* in our model) emerged as a strong predictor of sagebrush cover change (Fig. 2).
Despite previous evidence that spring temperature strongly regulates population- and individual-level sagebrush performance [@Schlaepfer2012; @Schlaepfer2014], we found little evidence for a temperature effect (Fig. 2).
Our estimates for temperature effects broadly overlapped zero, but they did trend positive.
Thus, we did find some support for the idea that sagebrush will respond to positively to temperature increases at higher elevations.
As seen when comparing our temporal forecasts (Fig. 5) to climate projections (Fig. 2), it also appears that our forecasts were driven by the weak, but positive, response to temperature.
Although our estimates of climate effects were small, they are in agreement with previous studies, indicating our model does an adequate job of capturing the ecological process and the link between sagebrush growth and climate.

>>>>

\textsc{\small{Andrew T. Tredennick\footnote{Corresponding author: atredenn@gmail.com}\textsuperscript{1}, Mevin B. Hooten\textsuperscript{2,3,4}, Cameron L. Aldridge\textsuperscript{5}, Collin G. Homer\textsuperscript{6}, Andrew Kleinhesselink\textsuperscript{1}, and Peter B. Adler\textsuperscript{1}}}

@@

Abstract
--------
Population modeling inference has traditionally been limited to small scale observational studies.
Yet, global environmental changes, like climate change, play out at landscape and regional scales, creating a mismatch between available tools and real world problems.
New technology is generating data products on massive scales covering entire ecosystems and teams are post-processing these data products to turn them into useful ecological information like species-specific estimates of plant percent cover.
Unlike some ecological sub-disciplines, population modelers have yet to fully embrace the use of remotely-sensed data as estimates of population abundance.
Such reluctance likely stems from challenges in fitting statistically appropriate models that rigorously account for spatial autocorrelation in large datasets.
Here, we use modern, computationally efficient model reparameterizations to fit a dynamic population model to a remotely-sensed dataset.
We use a 28 year time series of sagebrush (*Artemisia* spp.) percent cover over a $2.5\times5$ km landscape in southwestern Wyoming to fit a spatio-temporal population model that includes climate covariates and spatial random effects.
We then use the model to forecast future population states (percent cover) under projected climate scenarios.
Our analysis suggests sagebrush will increase in cover at this high elevation site, primarily driven by a positive response to previous year's precipitation.
However, gains in cover are not homogeneous in space.
Such spatially explicit estimates of increases/decreases in plant abundance have, until now, been elusive.
The spatiotemporal model we describe can be applied to any spatially explicit time series of plant cover.
As such datasets become increasingly available and precise, we expect the methods we use here to become an appealing alternative to species distribution models that lack population dynamics and, generally, estimates of species abundance.

@@

Introduction
-------------
Forecasting the impacts of climate change on plant populations and communities is a central challenge of modern ecology [@Clark2001; @Petchey2015].
Population models are ideally suited for meeting such a challenge because they allow inference on population abundance and dynamics (year-to-year changes in abundance) in relation to climate drivers [@Hare2010; @Adler2012; @Ross2015].
However, inference from population models is typically limited to small spatial extents because the data required is difficult to collect across wide species ranges.
Indeed, almost every study of plant population dynamics relies on demographic observations recorded at the meter to sub-meter scale [see, e.g., @Salguero-Gomez2015].
Local-scale demographic data make building population projection models an easy task [@Ellner2006; @Rees2009; @Adler2012], but it is very difficult to extrapolate small-scale studies to large spatial extents with any certainty because the data likely only represent a small subset of parameter space and environmental conditions [@Freckleton2011; @Queenborough2011].
Thus, the real challenge is not to simply make population forecasts, but to do so at spatial scales relevant to policy and management decisions [@Queenborough2011, Tredennick and Adler in review].

Species distribution models (SDMs), on the other hand, are easily applied at landscape to continental scales [e.g., @Maiorano2013; @Clark2014].
In a climate change context, SDMs typically rely on static associations between contemporary climate and a species' distribution or, more rarely, abundance to project future distribution or abundance [@Elith2009].
Thus, interannual dynamics that reflect population responses to yearly climate variations are not represented in SDMs.
Furthermore, since SDMs typically rely on occurence data, their projections of habitat suitability or probability of occurence provide little information on the future states of populations in the core of their range -- in areas where a species currently exists and is expected to persist in the future.
Despite these limitations, SDMs remain one of the few ways to make forecasts of species responses to climate change at large spatial extents.

The ideal tool for forecasting future population states under climate change would be a large-scale dynamic population model [@Schurr2012a; @Merow2014].
Unfortunately, such methods have been elusive due to the lack of time series data at large spatial extents and the statistical machinary needed to fit high-dimensional spatial models.
Fortunately, two recent technological advances put us in the position to develop large-scale population models.
First, remotely-sensed (RS) time series of species-specific plant cover can provide data on interannual population dynamics that can be correlated with yearly climate fluctuations at large spatial extents.
RS time series have revolutionized models of ecosystem-level processes [e.g., @Running2004] and have been used to detect long term trends in plant population abundance [e.g., @Homer2015], but they have yet to be used to drive a dynamic population model.
Second, animal population modelers have developed efficient dimension reduction and reparameterization techniques to fit high-dimension spatiotemporal models [see @Conn2015 for a review]. 
However, these methods have yet to be used in a dynamic setting.

Building on these technological advances, we develop a statistical approach for modeling plant populations over large spatial extents using a process model motivated by Gompertz density-dependent population growth.
Our approach leverages a unique remotely-sensed time series of *Artemisia* (sagebrush) species [@Homer2012] to fit a spatio-temporal model of plant population dynamics at the landscape scale.
We use advanced dimension reduction techniques borrowed from the animal modeling literature [@Conn2015] to rigorously account for spatial autocorrelation and to produce spatially-explicit estimates of sagebrush percent cover.
We then use the model to forecast the impacts of climate change on sagebrush percent cover.

Sagebrush ecosystems offer an ideal testing ground for our new modeling approach: sagebrush species are widely distributed [@Kuchler1964], sagebrush species are sensitive to climate [@Dalgleish2011; @Schlaepfer2014; @Schlaepfer2014a], and forecasts of future sagebrush ecosystems are in high demand due to the precarious conservation status of the greater sage-grouse (*Centrocercus urophasianus*) [@Arnett2015].
Species distribution models typically predict that much of the area occupied by sagebrush ecosystems today will become unsuitable for sagebrush due to climate change, resulting in a dramatic loss overall in the extent of sagebrush habitat by the end of this century [@Shafer2001; @Neilson2005; @Bradley2010; @Schlaepfer2012; @Still2015].
Sagebrush is thought to depend on winter snow and rain that recharges deeper soil layers [@Pechanec1937; @Schlaepfer2011; @Germino2014].
Models suggest that warming could drive sagebrush losses at low elevations and southern parts of its distribution by making snowmelt earlier and increasing evaporation, leading to soil water deficits in the spring [@Schlaepfer2012; @Schlaepfer2014].
In contrast, at higher elevations and colder regions, warming and earlier snowmelt could lengthen the growing season and increase sagebrush occurrence [@Schlaepfer2012; @Schlaepfer2014].
Thus, the response of sagebrush to climate will likely vary in space depending on elevation, among other factors.
Direct observations of individual plants and cover on experimental plots tend to agree with these models:  growth tends to respond negatively to spring and summer temperatures [@Miglia2005; @Poore2009; @Apodaca2013] except at higher elevations where earlier snowmelt may allow for a longer growing season and an increase in sagebrush growth and cover [@Perfors2003; @Harte2015].

@@

Materials and Methods
---------------------
### Data
#### Remotely-sensed time series
To demonstrate our modeling approach we use a subset of a remotely-sensed time series of sagebrush (\emph{Artemisia} spp.) shrub percent cover in Wyoming[^link0] [@Homer2012].
As part of a separate study, sagebrush percent cover was estimated using a regression tree to relate ground reflectances retrieved by three sources of optical imagery (QuickBird, Landsat, and AWiFS) to 1,780 field observations of sagebrush cover distributed across Wyoming.
The regression tree model was further validated using another 500 field observations. For Wyoming sagebrush, the model achieved an $R^{2}$ = 0.65.
To hind-cast sagebrush cover the regression tree model was applied to historical remote sensing images. 
This resulted in an annual time series of sagebrush cover at 30 meter resolution from 1984 to 2011.
In this dataset, values represent the percentage of a 30 $\times$ 30 meter pixel covered by sagebrush.
In our study, we focused on a 5,070 $\times$ 2,430 meter subset totaling 13,689 30 $\times$ 30 meter pixels each year (Figure 1).
Thus, the full dataset contains 369,603 observations consisting of 27 year-to-year transitions (27 years $\times$ 13,689 pixels).

@@

To forecast future population states, we required future projections of climate for our study area.
We used the most recent climate projections used by the Intergovernmental Panel on Climate Change (IPCC), the Coupled Model Intercomparison Project 5 (CMIP5; *available online*)[^link2].
The CMIP5 provides projections from a suite of global circulations models (GCMs), and we used projections from 18 GCMs (Table A1) that produced weather projections for all three ``Representative Concentration Pathways'': RCP 4.5, RCP 6.0, and RCP 8.5 (*described online*)[^link3].
The three RCPs correspond to stabilization of radiative forcing before 2100, after 2100, and increasing greenhouse gas emissions, respectively.
To simulate equilibrium sagebrush cover under projected future climate (see below), we averaged all model projections for the centroid of our study area over the time period of 2050 to 2100.
We did the same for the time period of 1950 to 2000 so we could calculate projected changes in climate.
To keep the temporal variability from observed climate consistent for climate change scenarios, we adjusted observed climate according to projected changes.
For temperature, we subtracted the average historic (1950-2000) temperature from the average projected (2050-2100) temperature. 
For precipitation, we calculated the proportional change: we subtracted average projected (2050-2100) precipitation from average historic (1950-2000) precipitation, and then divided by average historic (1950-2000) precipitation. 
We calculate precipitation and temperature changes for RCP 4.5, RCP 6.0, and RCP 8.5 scenarios.
We implemented projected temperature and precipitation changes in the model by perturbing observed annual values by the values in Table 1.

@@

In another set of simulations (see below), we used yearly GCM projections from 2012 to 2098 to make temporally-explicit forecasts starting from the end point of the remotely sensed sagebrush cover data (ends in 2011). We aggregated daily GCM output for each model and RCP scenario into the yearly climate covariates used to fit our model.

@@

\vspace{-4em}
\begin{align}
\text{log}(\mu_{i,t}) &= \underbrace{\beta_{0,t} + \beta_{1}y_{i,t-1}}_\text{temporal + dens. dep} + \underbrace{\textbf{x}_{t}'\boldsymbol{\gamma}}_\text{climate} + \underbrace{\eta_{i}}_\text{spatial}.
\end{align}
\vspace{-3em}

Our model of percent cover change includes a density-dependent effect of log-transformed cover in the previous year ($y_{i,t-1}$), climate effects ($\textbf{x}_t$), and a spatial random effect ($\eta$) for each pixel *i*.
Climate effects were standardized [$(x_{i}-\bar{x}) / \sigma(x)$] to improve convergence during the model fitting stage and to allow for easier prior specification.
The intercept, $\beta_{0,t}$, was allowed to vary through time so that climate variables were not over-fit.
We used a Poisson likelihood because integer percent cover values in the sagebrush data product can be considered a form of count data.
We also evaluated a negative binomial model, but found little evidence for overdispersion beyond what our model was already accomodating via the spatial random effects ($\boldsymbol{\eta}$). There was no evidence of zero-inflation in our data, but see below (*Accomodating zeros*) for how we dealt with the small number of zero percent cover observations.

Fitting the model with a spatial random effect ($\eta$) is computationally demanding when a large number of observations are considered.
This is due to the required calculations of the resulting covariance matrices, which increase as a cubic function of the number of locations [@Wikle2010].
To overcome these computational constraints, we employed a dimension reduction strategy to reduce the number of parameters that need to be estimated to account for spatial variation.
Reducing the number of effective spatial random effect parameters speeds up estimation by reducing the size of the covariance matrices that need to be inverted at each MCMC iteration.
Fitting models that appropriately account for spatial autocorrelation over large spatial extents would not be feasible without these modern techniques.
We follow a common dimension reduction strategy, and express the high dimensional spatial random effect, $\boldsymbol{\eta}$, as the product of an expansion matrix, \textbf{K}, and a smaller parameter vector, $\boldsymbol{\alpha}$ [e.g., @Hooten2003; @Hooten2007a; @Conn2015].
We can then define the spatial effect as

@@

\vspace{-4em}
\begin{align}
w_{s,m} &= \text{exp}\left(\frac{-d_{s,m}}{\sigma}\right)
\end{align}
\vspace{-4em}

where $d_{s,m}$ is the Euclidean distance between the centroid of sample cell \emph{s} and the location of knot \emph{m}, and $\sigma$ is the kernel bandwidth.
It is possible through exhaustive model selection and fitting to determine the optimal form of the kernel and to estimate optimal values for $\sigma$ [@Higdon2002; @Hooten2015].
However, given the relative size of our dataset and computational limitations, we defined kernels around 121 knots (Fig. B2) whose nearest neighbor distances are approximately equal to the range of spatial dependence in residuals from a simple GLM fit without climate covariates and the spatial random effect (Appendix B). The Bayesian posterior distribution can be expressed as

@@

### Computing
####Fitting the model
We fit the model in R [@R2013] using the 'No-U-Turn' Hamiltonian Monte Carlo sampler in Stan [@stan2014] and the RStan package [@rstan2014].
We obtained posterior distributions of all model parmaters from three MCMC chains comprised of 1,000 iterations each, after discarding an initial 1,000 iterations as burn in.
Such short MCMC chains may seem unusual to those more familiar with other MCMC samplers (e.g., JAGS or WinBUGS).
However, the Stan algorithm is extremely efficient, meaning that, compared to other samplers, fewer iterations are required to achieve convergence.
Each MCMC chain was initialized with unique parameter values and the model was fit in parallel using the Utah State University High-Performance Computing facility.
Model fitting required five days.
We assessed convergence visually and calculated scale reduction factors (Appendix C, $\hat{R}$ < 1.01 for all paramters) [@Gelman1992; @Gelman2009].

####Simulating the process
We performed four sets of simulations: (1) to compare observed and simulated equilibrium cover, (2) to compare observed and simulated year- and location-specific cover, (3) to forecast future equilibrium population states under projected climate change, and (4) to make temporally-explicit forecasts of sagebrush cover starting the final year of our observations and ending in year 2098. 
Using the posterior distribution of model parameters, we simulated a matrix of pixels equal to the size of the study area (13,689 pixels or matrix elements).
For simulations (1) and (3) we initialized all pixels with arbitrarily low cover (1%) and then projected the model forward by randomly drawing climate covariates from the observed climate time series (for 1) or a pertubed climate time series (for 3).
We ran equilibrium simulations (1 and 3) for 600 time steps and then compared the output across simulations, after discarding and initial 100 time steps.
For simulation (2), we initialized each pixel with its actual value for time *t* and cell *s* and projected the model forward one time step and compared the one-step forecast with the observed value.
For simulation (4), we initialized each pixel with the final observed value in 2011 and then projected the model forward based on GCM yearly weather projections.
We used the posterior mean of each parameter for all simulations, except for (4) where we ran 50 simulations with unique sets of parameters from the MCMC chains.
The random year effects were not used in any simulation.

@@

When we simulated the pixel-based population model based on observed climate it was able to reproduce average observed percent cover in space (Fig. 4A,B).
Our model shows a tendency to underpredict perennially-low percent cover pixels (Fig. 4C), but does a better job at predicting high cover pixels.
Nonetheless, observations and predictions are qualitatively similar (Fig. 4A, B).
Point predictions are most confident, though slightly biased, in low percent cover pixels (Fig. 4D).
The model is also able to adequately reproduce observed dynamics when we make one-step-ahead predictions based on observed climate and cover in the previous year for each pixel.
When we made these in-sample, one-step-ahead forecasts, the model achieved an RMSE = 4.31, in units of percent cover.
The correlation between observations and predictions was 0.62.

When we apply the fitted model to IPCC climate change scenarios, the model predicts gains in sagebrush percent cover, on average (Figs. 5, 6).
The spatial effect remains strong enough in low cover regions to counteract the amplifying effect of projected precipitation increases (Fig. 5).
Thus, our model predicts an increase in sagebrush cover heterogeneity because projected cover increases are smaller in low cover pixels than in high cover pixels. (Fig. 5).
For the temporally-explicit forecasts, we show spatially-averaged values and the associated uncertainty due to variability in GCM projections, variability in model parameters, and uncertainty in our process model (Fig. 6).
Based on our model and GCM projections, we forecast an average increase in sagebrush cover at our study area, but a decrease is not outside the realm of possibility (shaded regions in Fig. 6).
This result reflects the positive effect of precipitation on sagebrush cover change (Fig. 3).

Discussion
-----------
Landscape-scale population models for plant species have been elusive and challenging to implement [@Schurr2012a; @Merow2014].
Yet, the need to forecast population responses to climate change over large spatial extents is clear, as demonstrated by the wide application of species distribution models [see @Clark2014].
We introduced a new statistical approach to modeling plant populations that overcomes two major hurdles for spatially-explicit population models.
First, we used species-specific remote sensing data as a response variable.
Thus, we had a spatially-continuous data product of sagebrush abundance over a large spatial extent, meaning we could fit a model at the same spatial resolution and extent as the data. 
In so doing we, overcame a major limitation of most population models that are fit using detailed data from a single location.
Second, borrowing from the vast literature on spatio-temporal modeling of animal abundance [e.g., @Conn2015], we employed a dimension reduction strategy to rigorously account for spatial autocorrelation.
Accounting for spatial autocorrelation allows for statistically rigorous inference.
While important statistically, in terms of ecological inference, we can be confident in our estimates of climate effects on sagebrush cover change in our study region.
It also provided a way to obtain spatially-explicit predictions at a resolution below that of the climate covariates (Figs. 4,5).

The amount of precipitation in the previous year (*pptLag* in our model) emerged as a strong predictor of sagebrush cover change (Fig. 2).
Despite previous evidence that spring temperature strongly regulates population- and individual-level sagebrush performance [@Schlaepfer2012; @Schlaepfer2014], we found little evidence for a temperature effect (Fig. 2).
Our estimates for temperature effects broadly overlapped zero, but they did trend positive.
Thus, we did find some support for the idea that sagebrush will respond to positively to temperature increases colder parts of their range.
As seen when comparing our temporally-explicit forecasts (Fig. 6) to climate projections (Fig. 2), it also appears that our forecasts were driven by the weak, but positive, response to temperature.
Although our estimates of climate effects were small, they are in agreement with previous studies, indicating our model does an adequate job of capturing the ecological process and the link between sagebrush growth and climate.
